Diffusive Homeostasis in a Self-Organizing Recurrent Neural Network \\ \begin{large} Spatially dependent Interaction as a Determinant of Neural Activity and Plasticity \end{large}

### 2016-12-04 draft.tex
.

Returning to the question of how to describe NO synthesis in a simplified form, we note that according to equation \eqref{nNOS_single_sol_approx2}, a single spike causes the release of NO by an amount of $\frac{{Ca^{2+}_{spike}}^3 \tau_{Ca^{2+}}ln(2)}{3}$, which makes the mean rate of NO production over time simply $<nNOS> = <r> \cdot \frac{{Ca^{2+}_{spike}}^3 \tau_{Ca^{2+}}ln(2)}{3}$. More specific, if the average interspike intervals of the underlying spike generation were short compared to the decay constant $\tau_{nNOS}$, one could show that \eqref{nNOS_single_sol_approx2} approaches gaussian noise with the aforementioned mean $\mu_{nNOS}(<r>) =  <r> \cdot \frac{{Ca^{2+}_{spike}}^3 \tau_{Ca^{2+}}ln(2)}{3}$ and a variance of $\sigma^2_{nNOS}(<r>) = <r> \frac{{Ca^{2+}_{spike}}^6 {\tau_{Ca^{2+}}}^2 ln^2(2)}{18\tau_{nNOS}}$. However, for a single cell spiking at a mean firing rate of 3 Hz with approximately poissonian statistics, Figure \ref{nNOS_approx_plot} indicates that this approximation is not valid. Furthermore, the condition of interspike intervals to be short compared to the decay constant inherently contradicts the necessary assumptions that were taken for the previous approximations to hold.

To overcome this contradiction, we introduce an idealized picture of diffusive homeostasis, where diffusion across the tissue is instantaneous. This simplification results in a single level of NO-concentration for all neurons, only being modified over time by the sum of all neurons' NO-syntheses and the decay term $-\lambda NO$. To account for the spread of NO, $nNOS(t)$ must be divided by the area $L^2$ of the tissue. The idealized version of \eqref{NO_dyn} then reads
\begin{equation}
\dot{NO}(t)=-\lambda NO + \frac{\sum_{i} nNOS^i(t)}{L^2}\;.
\label{NO_dyn_approx}
\end{equation}
In this form, it becomes clear that we can now describe the \emph{total} influx of NO (the sum in \eqref{NO_dyn_approx}) as the sum of a mean and a random fluctuation: While spiking events on individual cells remain sufficiently separated to justify equation \eqref{nNOS_single_sol_approx2}, the sum of all spike events from the excitatory population results in an effective poisson process with a mean rate of $3 Hz \cdot N_{exc.} = \mathrm{1200\,Hz}$, given the assumption that cells are spiking weakly correlated. Furthermore, we argue that a total rate of $\mathrm{1200\,Hz}$ is sufficiently large to express this rate as a continuous function of time, $r_{total} = N_{exc.}\cdot r_{exc.pop.}(t)$, at least on the time scale of the oscillations to be studied (see  Figure \ref{rate_th_close}). Thus, equation \eqref{NO_dyn_approx} becomes
\begin{equation}
\dot{NO}(t)=-\lambda NO + \frac{\mu_{nNOS}(N_{exc.}r_{exc.pop.}(t)) + \sigma_{nNOS}(N_{exc.}r_{exc.pop.}(t)) \xi (t)}{L^2}\;.
\label{NO_dyn_approx2}
\end{equation}
As a next step, we attempted to simplify the dynamics of threshold adaption and its effect on the population rate. Since equation \eqref{NO_dyn_approx2} contains a single variable $r_{exc.pop.}(t)$ for the mean excitatory activity, we wanted to find an appropriate description containing the mean threshold $V_{t,pop.}(t) = <V_t(t)>$. As stated earlier, Figure \ref{rate_th_close} suggests an immediate and approximately linear relation between excitatory population rate and the mean of thresholds within the excitatory population. Figure \ref{thresh_r_linfit} shows a linear fit (being of the form $\alpha x + \beta$) of these two quantities plotted against each other. Apart from rarely appearing high values of $V_{t,pop.}$, we found a good fit to the linear model. However, when we color coded the slope of $V_{t,pop.}(t)$, one could see a separation of the rising phase and falling phase within the cloud of data points. This indicates a small (due to the "flatness" of the shape) but noticeable delay between threshold and firing rate, the latter following the former. We implemented this in a similar way as in Equation \eqref{2d-exc_inh_rate_1}. The population rate relaxates towards a value given by the fitted linear dependence on $V_{t,pop.}$ linear fit: $\dot{r_{exc.pop.}}(t) = (-r_{exc.pop.} + r_{IP} + \alpha(V_{t,pop.}(t)-V_{t,0}))/\tau_r$, where $r_{IP}$ is given in table \ref{Params_IP} as the mean excitatory target firing rate and $V_{t,0} \equiv (r_{IP}-\beta)/\alpha$ is the mean threshold corresponding to the target rate. We expect $\tau_r$ to be in the same order of magnitude as the membrane time constant $\tau_m$. However, $r_{exc.pop.}$ represents the whole excitatory population. Taking conduction delay and interactions with the inhibitory population into account, $\tau_r$ thus should be slightly larger.

>>>>

Returning to the question of how to describe NO synthesis in a simplified form, we note that according to equation \eqref{nNOS_single_sol_approx2}, a single spike causes the release of NO by an amount of $\frac{{Ca^{2+}_{spike}}^3 \tau_{Ca^{2+}}ln(2)}{3}$, which makes the mean rate of NO production over time simply $<nNOS> = <r> \cdot \frac{{Ca^{2+}_{spike}}^3 \tau_{Ca^{2+}}ln(2)}{3}$. However, for two reasons it is not sufficient to simply propose $nNOS_i(t) = r_i(t) \cdot \frac{{Ca^{2+}_{spike}}^3 \tau_{Ca^{2+}}ln(2)}{3}$. First, a single cell fires in the range of $\mathrm{3\,Hz}$, which is - at least on the time scale of the observed oscillations - not enough to assign an "instantaneous" firing rate. Second, the exponential kernel of nNOS leads to a certain amount of delay between the spiking activity and the rate of NO synthesis, see Figure \ref{rate_nNOS_lag}.
\begin{figure}
\begin{center}
\includegraphics[width=0.7\textwidth]{../../plots/diff_hom/rate_nNOS_close.png}
\end{center}
\caption{Average NO synthesis (red) lags behind the mean firing rate of the excitatory population (blue) by $\mathrm{\approx \, 0.1 \,s}$.}
\label{rate_nNOS_lag}
\end{figure}
To deal with this, we introduce an idealized picture of diffusive homeostasis, where diffusion across the tissue is instantaneous. This simplification results in a single level of NO-concentration for all neurons, only being modified over time by the sum of all neurons' NO-syntheses $nNOS_{total}(t) \equiv \sum_{i} nNOS^i(t)$ and the decay term $-\lambda NO$. To account for the spread of NO, $nNOS_{total}$ must be divided by the area $L^2$ of the tissue. Furthermore, we introduce a random fluctuation term $\sigma_{NO} \xi(t)$ that accounts for local, momentary deviations from the mean. The idealized version of \eqref{NO_dyn} then reads
\begin{equation}
\dot{NO}(t)=-\lambda NO + \frac{nNOS_{total}(t)}{L^2} + \sigma_{NO} \xi(t)\;.
\label{NO_dyn_approx}
\end{equation}
The amount of noise in the system is a hard-to-predict quantity since it depends on the spatial structure on the diffusive lattice, giving neurons that are further separated less impact on each other. We tried to estimate the noise level based on the same "instant spread"-simplification made in Equation \eqref{NO_dyn_approx} to describe the overall NO-dynamics. In the final result, this led to an underestimation of the total energy within the power spectrum. However, as long as we assume the noise in the system to be approximately white, noise level only acts as a global scaling factor of the shape of the power spectrum and was thus left open as a model fitting parameter.

From Equation \eqref{NO_dyn_approx} it is clear why we can now describe the \emph{total} influx of NO as the sum of a mean and a random fluctuation: While spiking events on individual cells remain sufficiently separated to justify equation \eqref{nNOS_single_sol_approx2}, the sum of all spike events from the excitatory population results in an effective poisson process with a mean rate of $3 Hz \cdot N_{exc.} = \mathrm{1200\,Hz}$, given the assumption that cells are spiking weakly correlated. Furthermore, we argue that a total rate of $\mathrm{1200\,Hz}$ is sufficiently large to express this rate as a continuous function of time, $r_{total} = N_{exc.}\cdot r_{exc.pop.}(t)$, at least on the time scale of the oscillations to be studied (see  Figure \ref{rate_th_close}). Still though, we need to address the fact that $nNOS_{total}(t)$ lags behind $r_{total}$. We do so by a dynamical instead of a purely functional relationship:
\begin{equation}
\dot{nNOS_{total}}(t) = \frac{r_{exc.pop.}\cdot N_{exc.}\frac{{Ca^{2+}_{spike}}^3 \tau_{Ca^{2+}}ln(2)}{3} - nNOS_{total}}{\tau_{nNOS}}
\label{nNOS_total_dyn}
\end{equation} 
where $\frac{{Ca^{2+}_{spike}}^3 \tau_{Ca^{2+}}ln(2)}{3}$ is the integral from Equation \eqref{nNOS_single_sol_approx1}.

As a next step, we attempted to simplify the dynamics of threshold adaption and its effect on the population rate. Corresponding to the reduction to a single variable $r_{exc.pop.}(t)$ for the mean excitatory activity in equation \eqref{nNOS_total_dyn}, we wanted to find an appropriate description containing only the mean threshold $V_{t,pop.}(t) = <V_t(t)>$. As stated earlier, Figure \ref{rate_th_close} suggests an immediate and approximately linear relation between excitatory population rate and the mean of thresholds within the excitatory population. Figure \ref{thresh_r_linfit} shows a linear fit (being of the form $\alpha x + \beta$) of these two quantities plotted against each other. Apart from rarely appearing high values of $V_{t,pop.}$, we found a good fit to the linear model. However, when we color coded the slope of $V_{t,pop.}(t)$, one could see a separation of the rising phase and falling phase within the cloud of data points. This indicates a small (due to the "flatness" of the shape) but noticeable delay between threshold and firing rate, the latter following the former. We implemented this in a similar way as in Equation \eqref{2d-exc_inh_rate_1}. The population rate relaxates towards a value given by the fitted linear dependence on $V_{t,pop.}$ linear fit: $\dot{r_{exc.pop.}}(t) = (-r_{exc.pop.} + r_{IP} + \alpha(V_{t,pop.}(t)-V_{t,0}))/\tau_r$, where $r_{IP}$ is given in table \ref{Params_IP} as the mean excitatory target firing rate and $V_{t,0} \equiv (r_{IP}-\beta)/\alpha$ is the mean threshold corresponding to the target rate. We expect $\tau_r$ to be in the same order of magnitude as the membrane time constant $\tau_m$. However, $r_{exc.pop.}$ represents the whole excitatory population. Taking conduction delay and interactions with the inhibitory population into account, $\tau_r$ thus should be slightly larger.
